Bootstrapping Multi-atlas Segmentation Using Multiple Automatically Generated Templates for the Segmentation of the Whole Hippocampus and Subfields

### 2013-09-03 paper.Rnw
Fixed up uses of newcommands

\textbf{Introduction:} 
Advances in image segmentation of magnetic resonance images (MRI) have
demonstrated that multi-atlas approaches improve segmentation accuracy and
precision over regular atlas-based approaches.  These approaches often rely on
a large number of such manually segmented atlases (e.g.  30-80) that take
significant time and expertise to produce. We present an algorithm, \mb (\mb
({\textbf M}ultiple {\textbf A}utomatically {\textbf Ge}nerated {\textbf
T}emplates), for the automatic segmentation of the hippocampus that minimizes
the number of atlases needed while still achieving similar accuracy to
multi-atlas approaches. Thus, our method acts as an accurate multi-atlas
approach when using special, hard-to-define atlases that are laborious to
construct. \\
\textbf{Method:} \mb works by propagating atlas segmentations to a template library,
formed from a subset of target images, via transformations estimated by non-linear
image registration. The resultant segmentations are then propagated to each
target image and fused using a label fusion method. We conduct a 10-fold Monte
Carlo cross-validation of whole hippocampal segmentation on a pool of 69 ADNI
subjects over a range of parameter settings, and a leave-one-out
cross-validation (LOOCV) of hippocampal subfield segmentation using five
high-resolution atlases. Two final experiments assess \mb when applied to
first episode psychosis and Alzheimer's disease populations, and \mb
segmentations are compared with existing automated methods (\fsl, \freesurfer,
\maper) and biases are explored.\\
\textbf{Results:} Using nine atlases and 19 template images, \mb achieves a
mean Dice's Similarity Coefficient (DSC) of 0.84 over 10-folds of Monte Carlo
cross-validation on 69 subjects relative to semi-automated \gs segmentations,
and shows significantly lower variability in DSC than multi-atlas segmentation.
In a LOOCV, \mb reproduces the volumes of the cornu ammonis (CA) 1; CA4/dentate
gyrus (DG); and strata radiatum (SR), strata lacunosum (SL), and strata
moleculare (SM) hippocampal subregions with a percent error in volume that is
at or lower than that produced by image resampling. \mb produces hippocampal
volumes in a first episode psychosis patient population that are highly
correlated with expert manual segmentation volumes (Pearson $r = 0.877, t =
16.244, p < 0.001$). Compared to \fsl and \freesurfer, \mb shows much smaller
fixed volume bias (within $250 mm^3$ on average) to \gs segmentations available
from ADNI, as well as a conservative, rather than exaggerated, proportional
volume bias.\\ \textbf{Conclusion:} We demonstrate that \mb produces accurate
hippocampal segmentations using only 5 atlases over different hippocampal
definitions, disease populations, and acquisition types, as well as showing
that accurate identification of the hippocampal subfields is possible.
\end{abstract}

@@

One methodology that can be used to mitigate these sources of errors involves
the use of multiple manually segmented atlases and probabilistic segmentation
techniques, such as those found in the \freesurfer package \citep{Fischl2002}.
\freesurfer uses a probabilistic atlas of anatomical and tissue classes along
with spatial constraints for class labels encoded using a Markov random field
model to segment the entire brain.

@@

Of special relevance to the present work is the LEAP algorithm (Learning
Embeddings for Atlas Propagation; \citet{Wolz2010}) because of its focus on
performing multi-atlas segmentation with a limited number of input atlases.  The
LEAP algorithm is a clever modification to the basic multi-atlas strategy in
which an atlas library is grown, beginning with a set of manually labelled
atlases, by successively incorporating unlabelled target images once they
themselves have been labelled using multi-atlas techniques. The sequence in
which target images are labelled is chosen so that the similarity between the
atlas images and the target images is minimised at each step, effectively
allowing for deformations between very dissimilar images to be broken up into
sequences of smaller deformations. Although \citet{Wolz2010} begin with an atlas
library of 30 MR images, this method could theoretically work using a much
smaller atlas library.  In their validation, LEAP was used to segment the whole
hippocampus in the ADNI-1 baseline dataset, achieving a mean Dice score of 0.85
against \gs segmentations.

Also of interest to this manuscript are methods that attempt to define
hippocampal subfields using standard T1-weighted data.  To the best of our
knowledge, there are only two automated segmentation algorithms that attempt
this problem. The first is included with the \freesurfer package
\citep{VanLeemput2009}. This work is limited as it omits the tail of the
hippocampus and the segmentation protocol has yet to be fully validated.
Nonetheless, it demonstrates that the applicability of hippocampal subfield
segmentation using data from 10 subjects. In the second method,
\citet{Yushkevich2009} hippocampal subfields were acquired and labelled on
high-resolution MRI data from post-mortem medial temporal lobe samples. Using
nonlinear registration guided by manually derived hippocampus masks and
specific landmarks, the authors demonstrated accurate parcellation of
hippocampal subfields in unlabelled MRI volumes.

In this paper we have describe a thorough validation of the \mb algorithm for
segmentation of the hippocampus and its subfields. First, we addressed the feasibility and
accuracy of whole hippocampus segmentation with a limited number of input
atlases \citep{MallarChakravarty2012} by performing a multi-fold experiment
using a subset of the Alzheimer's Disease Neuroimaging Initiative (ADNI)
dataset over a range of parameters.  We then performed a leave-one-out
validation to determine if hippocampal subfields can be accurately identified
using our multi-atlas framework.  To ensure that we have not over-fit our
parameters to the aging or neurodegenerative brain, we also applied our
segmentations to a dataset of individuals suffering from first episode
psychosis.  Finally, validated our algorithm using all of the data available in
the \adnidataset sample and compare our segmentations to the other
segmentations available through the ADNI informatics portal.

@@

\subsection{The \mb Algorithm}
In this paper, we use the term {\em label} to mean any segmentation (manual or
derived) of an MR image. {\em Label propagation} is the process by which two
images are registered and the resulting transformation is applied to the labels
from one image to bring them into alignment with the other image. We use the
term {\em atlas} to mean a manually segmented image, and the term {\em template}
to mean an automatically segmented image (i.e.  via label propagation). The
terms {\em atlas library} and {\em template library} describe any set of such
images. Additionally, we use the term {\em target} to refer to an unlabelled
image that is undergoing segmentation.

@@

\mb ({\textbf M}ultiple {\textbf A}utomatically {\textbf Ge}nerated {\textbf
T}emplates) bootstraps the creation of a large template library given a limited
input atlas library, and then uses the template library in basic multi-atlas
segmentation.  Images for the template library are selected from a set of input
target images, either arbitrarily or so as to reflect the neuroanatomy or
demographics of the target set as a whole (for instance, by sampling equally
from cases and controls).  The template library images are automatically
labelled by each of the atlases via label propagation.  Basic multi-atlas
segmentation is then conducted using the template library to segment the entire
set of target images (including the target images used in the construction of
the template library).  Since each template library image has multiple labels
(one from each atlas), the final number of labels to be fused for each target
may be quite large (i.e. \# of atlas $\times$ \# of templates).

Figure \ref{fig:MAGeT} illustrates the \mb algorithm graphically. Source code
for \mb can be found at \url{http://github.com/pipitone/MAGeTbrain}.

\begin{figure}
  \begin{centering}
    \includegraphics[width=\textwidth]{figure/MA-MAGeTBrain-Schematic}
  \end{centering}
  \caption{A schematic illustration of basic multi-atlas segmentation and \mb
  segmentation \label{fig:MAGeT}}
\end{figure}

@@

The following section describes for experiments conducted to assess the
segmentation quality of the \mb algorithm. The first two experiments assess the
validity of \mb using a cross-validation design. Experiment 1 investigates the
accuracy of \mb whole hippocampus segmentation over a wide range of parameter
settings. The results of this experiment enable us to choose the parameter
settings offering the best performance for use in subsequent experiments.
Experiment 2 assesses hippocampal subfield segmentation quality in a
leave-one-out cross-validation design. The last two experiments assess the
validity of the \mb algorithm when applied to different diseases: first episode
schizophrenia (Experiment 3), and Alzheimer's disease (Experiment 4).
Additionally, in Experiment 4, we compared \mb segmentations with those of
well-known automated methods and assessed segmentation bias.

@@

Monte Carlo Cross-Validation (MCCV) \citep{Shao1993} was performed using a pool
of images and SNT \gs segmentations from the ADNI dataset \citep{Hsu2002}. This
form of cross-validation allows us to rigorously validate a large number of
parameter settings of \mb (atlas and template library sizes, registration
algorithm, and label fusion method) and select the best parameters to use in
subsequent experiments.

\paragraph{\adnidataset dataset} 
Data used in the preparation of this article were obtained from the Alzheimer’s
Disease Neuroimaging Initiative (ADNI) database (adni.loni.ucla.edu). The ADNI
was launched in 2003 by the National Institute on Aging (NIA), the National
Institute of Biomedical Imaging and Bioengineering (NIBIB), the Food and Drug
Administration (FDA), private pharmaceutical companies and non-profit
organizations, as a \$60 million, 5-year public- private partnership. The primary
goal of ADNI has been to test whether serial magnetic resonance imaging (MRI),
positron emission tomography (PET), other biological markers, and clinical and
neuropsychological assessment can be combined to measure the progression of mild
cognitive impairment (MCI) and early Alzheimer’s disease (AD).  Determination of
sensitive and specific markers of very early AD progression is intended to aid
researchers and clinicians to develop new treatments and monitor their
effectiveness, as well as lessen the time and cost of clinical trials.

@@

69 1.5T images were arbitrarily selected from the baseline scans in the {\em
\adnidataset} standardized dataset. 23 subjects were chosen from each disease
category: cognitively normal (CN), mild cognitive impairment (MCI) and
Alzheimer's disease (AD). Demographics for this subset are shown in Table
\ref{tab:ADNI1-xvalidation-demographics}.  Semi-automated segmentations of the
left and right whole hippocampi are made available with a subset of ADNI images
\citep{Hsu2002}. These labels have been generated using the SNT tool from
Medtronic Surgical Navigation Technologies, Louisville, CO (see Supplementary
Materials for detailed discussion of the segmentation process used). Throughout
this paper, we use these segmentations as the \gs for comparison.

Clinical, demographic and pre-processed T1-weighted MRI were downloaded by the
authors from the ADNI database (adni.loni.ucla.edu) between March 2012 and
August 2012. The image dataset used was the "\adnidataset" standardized dataset
available from ADNI \footnote{
\url{http://adni.loni.ucla.edu/methods/mri-analysis/adni-standardized-data/}}
\citep{Wyman2012}. This image collection contains uniformly pre-processed images
which have been designated to be the "best" after quality control.  All images
were acquired using 1.5T scanners (General Electric Healthcare, Philips Medical
Systems or Siemens Medical Solutions) at multiple sites using the protocol
described in \citep{Jack2008}.  Representative 1.5T imaging parameters were TR =
2400ms, TI = 1000ms, TE = 3.5ms, flip angle = $8^{\circ}$, field of view = 240 x
240mm, a $192 \times 192 \times 166$ matrix ($x$, $y$, and $z$ directions)
yielding a voxel dimensions of $1.25mm \times 1.25mm \times 1.2mm$.

@@

In this experiment, our dataset consists of 69 1.5T images and corresponding \gs
segmentations. In each validation round, the dataset is partitioned into a
training set consisting of images and \gs segmentations used as an atlas
library, and a validation set consisting of the remaining images segmented by
both \mb and multi-atlas. The computed segmentations were compared to the \gs
segmentations (see below).

A total of ten validation rounds were performed on each subject in the dataset,
over each combination of parameter settings. The parameter settings
explored are: atlas library size (1-9), template library size (1-20),
registration method (\ants or \animal, described below), and label fusion method (majority vote,
cross-correlation weighted majority vote, and normalized mutual information
weighted majority vote, described below).  A total of $10 \times 69 \times 9
\times 20 \times 2 \times 3 = \num{7452000}$ validation rounds were conducted,
resulting in a total of $\num{1490400}$ segmentations analysed.

@@

\subparagraph{Automatic Normalization and Image Matching and Anatomical
Labeling (\animal)}

The \animal algorithm carries out image registration in two phases. In the
first, a 12-parameter linear transformation (3 translations, rotations, scales,
shears) is estimated between images using an algorithm that maximizes the
correlation between blurred MR intensities and gradient magnitude over the whole
brain \citep{Collins}. In the second phase, nonlinear registration is completed
using the \animal algorithm \citep{Collins1995}: an iterative procedure that
estimates a 3D deformation field between two MR images. At first, large
deformations are estimated using a blurred version of the input data. These larger
deformations are then input to subsequent steps where the fit is refined by
estimating smaller deformations on data blurred with a Gaussian kernel with a
smaller full width at half maximum (FWHM). The final transformation is a set of
local translations defined on a bed of equally spaced nodes that were estimated
through the optimization of the correlation coefficient.  For the purposes of
this work we used the regularization parameters optimized in
\citet{Robbins2004}, displayed in Table \ref{tab:ANIMAL-params}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ANIMAL parameters                    
%                                     
\begin{table*}[!tbp]
\scriptsize
\caption{\textbf{\animal registration parameters.} \label{tab:ANIMAL-params}}
\begin{center}
\begin{tabular}{l | c c c}
\hline 
Parameters        & Stage 1 & Stage 2 & Stage 3  \tabularnewline
\hline 
Model Blur (FWHM) &   8     &    8    &   4      \tabularnewline
Input Blur (FWHM) &   8     &    8    &   4      \tabularnewline
Iterations        &   30    &    30   &   10     \tabularnewline
Step              &   8x8x8 &  4x4x4  &   2x2x2  \tabularnewline
Sub-Lattice       &   6     &    6    &   6      \tabularnewline
Lattice Diameter  &24x24x24 &12x12x12 & 6x6x6    \tabularnewline
\hline
\end{tabular}
\end{center}
\end{table*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subparagraph{Automatic Normalization Tools (\ants)}

\ants is a diffeomorphic registration algorithm which provides great flexibility
over the choice of transformation model, objective function, and the consistency
of the final transformation \citep{Avants2008}. The transformation is estimated in a
hierarchical fashion where the MRI data is subsampled, allowing large
deformations to be estimated and successively refined at later hierarchical
stages (where the data is subsampled to a finer grid). The deformation field and
the objective function are regularized with a Gaussian kernel at each level of
the hierarchy. The \ants algorithm is freely available
\url{http://www.picsl.upenn.edu/ANTS/}. We used an implementation of the \ants
algorithm compatible with the MINC data format, mincANTS
\url{https://github.com/vfonov/mincANTS}.

We used the following command line when running \ants:
{\small
\begin{verbatim}
  mincANTS 3 -m PR[target_file.mnc,source_file.mnc,1,4] 
   --number-of-affine-iterations 10000x10000x10000x10000x10000 
   --affine-gradient-descent-option 0.5x0.95x1.e-4x1.e-4
   --use-Histogram-Matching --MI-option 32x16000
   -r Gauss[3,0] -t SyN[0.5] -i 100x100x100x20
   -o transformation.xfm
 \end{verbatim}
}
These settings were adapted from the "reasonable starting point" given in the
\ants manual 
\footnote{\url{https://sourceforge.net/projects/advants/files/Documentation/}}.

@@

  The {\tt xcorr\_vol} utility from the \animal toolkit is used to calculate the
  cross-correlation similarity measure.

@@

The automatically produces labels are compared to the \gs SNT labels provided by
ADNI. The segmentation accuracy reported for a parameter setting is the average
over ten validation rounds.

In order to investigate the performance of \mb in a real world setting in which 
only one size of atlas and template libraries are used, we explore the
variability in label agreement at fixed parameter settings under varying  
atlas and template library compositions. This is achieved by first computing the
standard deviation and variance of DSC scores in each block of ten validation
rounds per subject. The distribution of these statistics across all subjects is
then compared between \mb and multi-atlas using a Student's t-test. A
significant difference between distributions is taken to show either a larger or
smaller level of variability between methods.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%        Experiment 2: Hippocampal Subfield Cross-Validation          %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Experiment 2: Hippocampal Subfield Cross-Validation}
In this experiment, the accuracy of the \mb algorithm on hippocampal subregion
segmentation is tested using a leave-one-out cross-validation (LOOCV) design on
two image sets. The optimal parameter settings for \mb found in Experiment 1 are
used in this experiment.

@@

In part 2a of this experiment, we have modified the LOOCV experimental
design in order to compare \mb segmentation accuracy with the accuracy of image
resampling error. The five $0.3mm$-isotropic voxel Winterburn atlas images and
labels are used as the atlas library, and LOOCV is conducted using the five Winterburn
atlas images subsampled (using trilinear interpolation) to
0.9mm-isotropic voxel resolution (referred to as the {\em Subsampled} dataset)
as input subjects. Subsampling the target images allows us to assess \mb in a
typical segmentation scenario (high-resolution atlases and lower-resolution
subjects).  The template library is composed of all five subject images, plus
an additional set of 16 of healthy subjects 3T T1 images acquired separately
(Table \ref{tab:WAval-healthy-demographics}).  Images were acquired on a 3T GE
Discovery MR 750 system (General Electric, Milwaukee, WI) using an 8-channel
head coil with the enhanced fast gradient recalled echo 3-dimensional
acquisition protocol, EFGRE-BRAVO, with the following parameters: $TE/TR/TI =
3.0ms/6.7ms/650ms$, flip angle=$8^\circ$ , $FOV = 15.3cm$, slice
thickness$=0.9mm$, 170 in-plane steps for an approximate isotropic resolution
of 0.9mm dimension voxels.

>>>>

\textbf{Introduction:} 
Advances in image segmentation of magnetic resonance images (MRI) have
demonstrated that multi-atlas approaches improve segmentation accuracy and
precision over regular atlas-based approaches.  These approaches often rely on
a large number of such manually segmented atlases (e.g.  30-80) that take
significant time and expertise to produce. We present an algorithm, \mb{} (
({\textbf M}ultiple {\textbf A}utomatically {\textbf Ge}nerated {\textbf
T}emplates), for the automatic segmentation of the hippocampus that minimizes
the number of atlases needed while still achieving similar accuracy to
multi-atlas approaches. Thus, our method acts as an accurate multi-atlas
approach when using special, hard-to-define atlases that are laborious to
construct. \\
\textbf{Method:} \mb{} works by propagating atlas segmentations to a template library,
formed from a subset of target images, via transformations estimated by non-linear
image registration. The resultant segmentations are then propagated to each
target image and fused using a label fusion method. We conduct a 10-fold Monte
Carlo cross-validation of whole hippocampal segmentation on a pool of 69 ADNI
subjects over a range of parameter settings, and a leave-one-out
cross-validation (LOOCV) of hippocampal subfield segmentation using five
high-resolution atlases. Two final experiments assess \mb{} when applied to
first episode psychosis and Alzheimer's disease populations, and \mb{}
segmentations are compared with existing automated methods (\fsl{}, \freesurfer{},
\maper{}) and biases are explored.\\
\textbf{Results:} Using nine atlases and 19 template images, \mb{} achieves a
mean Dice's Similarity Coefficient (DSC) of 0.84 over 10-folds of Monte Carlo
cross-validation on 69 subjects relative to semi-automated \gs{} segmentations,
and shows significantly lower variability in DSC than multi-atlas segmentation.
In a LOOCV, \mb{} reproduces the volumes of the cornu ammonis (CA) 1; CA4/dentate
gyrus (DG); and strata radiatum (SR), strata lacunosum (SL), and strata
moleculare (SM) hippocampal subregions with a percent error in volume that is
at or lower than that produced by image resampling. \mb{} produces hippocampal
volumes in a first episode psychosis patient population that are highly
correlated with expert manual segmentation volumes (Pearson $r = 0.877, t =
16.244, p < 0.001$). Compared to \fsl{} and \freesurfer{}, \mb{} shows much smaller
fixed volume bias (within $250 mm^3$ on average) to \gs{} segmentations available
from ADNI, as well as a conservative, rather than exaggerated, proportional
volume bias.\\ \textbf{Conclusion:} We demonstrate that \mb{} produces accurate
hippocampal segmentations using only 5 atlases over different hippocampal
definitions, disease populations, and acquisition types, as well as showing
that accurate identification of the hippocampal subfields is possible.
\end{abstract}

@@

One methodology that can be used to mitigate these sources of errors involves
the use of multiple manually segmented atlases and probabilistic segmentation
techniques, such as those found in the \freesurfer{} package \citep{Fischl2002}.
\freesurfer{} uses a probabilistic atlas of anatomical and tissue classes along
with spatial constraints for class labels encoded using a Markov random field
model to segment the entire brain.

@@

Of special relevance to the present work is the LEAP algorithm (Learning
Embeddings for Atlas Propagation; \citet{Wolz2010}) because of its focus on
performing multi-atlas segmentation with a limited number of input atlases.  The
LEAP algorithm is a clever modification to the basic multi-atlas strategy in
which an atlas library is grown, beginning with a set of manually labelled
atlases, by successively incorporating unlabelled target images once they
themselves have been labelled using multi-atlas techniques. The sequence in
which target images are labelled is chosen so that the similarity between the
atlas images and the target images is minimised at each step, effectively
allowing for deformations between very dissimilar images to be broken up into
sequences of smaller deformations. Although \citet{Wolz2010} begin with an atlas
library of 30 MR images, this method could theoretically work using a much
smaller atlas library.  In their validation, LEAP was used to segment the whole
hippocampus in the ADNI-1 baseline dataset, achieving a mean Dice score of 0.85
against \gs{} segmentations.

Also of interest to this manuscript are methods that attempt to define
hippocampal subfields using standard T1-weighted data.  To the best of our
knowledge, there are only two automated segmentation algorithms that attempt
this problem. The first is included with the \freesurfer{} package
\citep{VanLeemput2009}. This work is limited as it omits the tail of the
hippocampus and the segmentation protocol has yet to be fully validated.
Nonetheless, it demonstrates that the applicability of hippocampal subfield
segmentation using data from 10 subjects. In the second method,
\citet{Yushkevich2009} hippocampal subfields were acquired and labelled on
high-resolution MRI data from post-mortem medial temporal lobe samples. Using
nonlinear registration guided by manually derived hippocampus masks and
specific landmarks, the authors demonstrated accurate parcellation of
hippocampal subfields in unlabelled MRI volumes.

In this paper we have describe a thorough validation of the \mb{} algorithm for
segmentation of the hippocampus and its subfields. First, we addressed the feasibility and
accuracy of whole hippocampus segmentation with a limited number of input
atlases \citep{MallarChakravarty2012} by performing a multi-fold experiment
using a subset of the Alzheimer's Disease Neuroimaging Initiative (ADNI)
dataset over a range of parameters.  We then performed a leave-one-out
validation to determine if hippocampal subfields can be accurately identified
using our multi-atlas framework.  To ensure that we have not over-fit our
parameters to the aging or neurodegenerative brain, we also applied our
segmentations to a dataset of individuals suffering from first episode
psychosis.  Finally, validated our algorithm using all of the data available in
the \adnidataset{} sample and compare our segmentations to the other
segmentations available through the ADNI informatics portal.

@@

\subsection{The \mb{} Algorithm}
In this paper, we use the term {\em label} to mean any segmentation (manual or
derived) of an MR image. {\em Label propagation} is the process by which two
images are registered and the resulting transformation is applied to the labels
from one image to bring them into alignment with the other image. We use the
term {\em atlas} to mean a manually segmented image, and the term {\em template}
to mean an automatically segmented image (i.e.  via label propagation). The
terms {\em atlas library} and {\em template library} describe any set of such
images. Additionally, we use the term {\em target} to refer to an unlabelled
image that is undergoing segmentation.

@@

\mb{} ({\textbf M}ultiple {\textbf A}utomatically {\textbf Ge}nerated {\textbf
T}emplates) bootstraps the creation of a large template library given a limited
input atlas library, and then uses the template library in basic multi-atlas
segmentation.  Images for the template library are selected from a set of input
target images, either arbitrarily or so as to reflect the neuroanatomy or
demographics of the target set as a whole (for instance, by sampling equally
from cases and controls).  The template library images are automatically
labelled by each of the atlases via label propagation.  Basic multi-atlas
segmentation is then conducted using the template library to segment the entire
set of target images (including the target images used in the construction of
the template library).  Since each template library image has multiple labels
(one from each atlas), the final number of labels to be fused for each target
may be quite large (i.e. \# of atlas $\times$ \# of templates).

Figure \ref{fig:MAGeT} illustrates the \mb{} algorithm graphically. Source code
for \mb{} can be found at \url{http://github.com/pipitone/MAGeTbrain}.

\begin{figure}
  \begin{centering}
    \includegraphics[width=\textwidth]{figure/MA-MAGeTBrain-Schematic}
  \end{centering}
  \caption{A schematic illustration of basic multi-atlas segmentation and \mb{}
  segmentation \label{fig:MAGeT}}
\end{figure}

@@

The following section describes for experiments conducted to assess the
segmentation quality of the \mb{} algorithm. The first two experiments assess the
validity of \mb{} using a cross-validation design. Experiment 1 investigates the
accuracy of \mb{} whole hippocampus segmentation over a wide range of parameter
settings. The results of this experiment enable us to choose the parameter
settings offering the best performance for use in subsequent experiments.
Experiment 2 assesses hippocampal subfield segmentation quality in a
leave-one-out cross-validation design. The last two experiments assess the
validity of the \mb{} algorithm when applied to different diseases: first episode
schizophrenia (Experiment 3), and Alzheimer's disease (Experiment 4).
Additionally, in Experiment 4, we compared \mb{} segmentations with those of
well-known automated methods and assessed segmentation bias.

@@

Monte Carlo Cross-Validation (MCCV) \citep{Shao1993} was performed using a pool
of images and SNT \gs{} segmentations from the ADNI dataset \citep{Hsu2002}. This
form of cross-validation allows us to rigorously validate a large number of
parameter settings of \mb{} (atlas and template library sizes, registration
algorithm, and label fusion method) and select the best parameters to use in
subsequent experiments.

\paragraph{\adnidataset{} dataset} 
Data used in the preparation of this article were obtained from the Alzheimer’s
Disease Neuroimaging Initiative (ADNI) database (adni.loni.ucla.edu). The ADNI
was launched in 2003 by the National Institute on Aging (NIA), the National
Institute of Biomedical Imaging and Bioengineering (NIBIB), the Food and Drug
Administration (FDA), private pharmaceutical companies and non-profit
organizations, as a \$60 million, 5-year public- private partnership. The primary
goal of ADNI has been to test whether serial magnetic resonance imaging (MRI),
positron emission tomography (PET), other biological markers, and clinical and
neuropsychological assessment can be combined to measure the progression of mild
cognitive impairment (MCI) and early Alzheimer’s disease (AD).  Determination of
sensitive and specific markers of very early AD progression is intended to aid
researchers and clinicians to develop new treatments and monitor their
effectiveness, as well as lessen the time and cost of clinical trials.

@@

69 1.5T images were arbitrarily selected from the baseline scans in the {\em
\adnidataset{}} standardized dataset. 23 subjects were chosen from each disease
category: cognitively normal (CN), mild cognitive impairment (MCI) and
Alzheimer's disease (AD). Demographics for this subset are shown in Table
\ref{tab:ADNI1-xvalidation-demographics}.  Semi-automated segmentations of the
left and right whole hippocampi are made available with a subset of ADNI images
\citep{Hsu2002}. These labels have been generated using the SNT tool from
Medtronic Surgical Navigation Technologies, Louisville, CO (see Supplementary
Materials for detailed discussion of the segmentation process used). Throughout
this paper, we use these segmentations as the \gs{} for comparison.

Clinical, demographic and pre-processed T1-weighted MRI were downloaded by the
authors from the ADNI database (adni.loni.ucla.edu) between March 2012 and
August 2012. The image dataset used was the "\adnidataset{}" standardized dataset
available from ADNI \footnote{
\url{http://adni.loni.ucla.edu/methods/mri-analysis/adni-standardized-data/}}
\citep{Wyman2012}. This image collection contains uniformly pre-processed images
which have been designated to be the "best" after quality control.  All images
were acquired using 1.5T scanners (General Electric Healthcare, Philips Medical
Systems or Siemens Medical Solutions) at multiple sites using the protocol
described in \citep{Jack2008}.  Representative 1.5T imaging parameters were TR =
2400ms, TI = 1000ms, TE = 3.5ms, flip angle = $8^{\circ}$, field of view = 240 x
240mm, a $192 \times 192 \times 166$ matrix ($x$, $y$, and $z$ directions)
yielding a voxel dimensions of $1.25mm \times 1.25mm \times 1.2mm$.

@@

In this experiment, our dataset consists of 69 1.5T images and corresponding \gs{}
segmentations. In each validation round, the dataset is partitioned into a
training set consisting of images and \gs{} segmentations used as an atlas
library, and a validation set consisting of the remaining images segmented by
both \mb{} and multi-atlas. The computed segmentations were compared to the \gs{}
segmentations (see below).

A total of ten validation rounds were performed on each subject in the dataset,
over each combination of parameter settings. The parameter settings
explored are: atlas library size (1-9), template library size (1-20),
registration method (\ants{} or \animal{}, described below), and label fusion method (majority vote,
cross-correlation weighted majority vote, and normalized mutual information
weighted majority vote, described below).  A total of $10 \times 69 \times 9
\times 20 \times 2 \times 3 = \num{7452000}$ validation rounds were conducted,
resulting in a total of $\num{1490400}$ segmentations analysed.

@@

\subparagraph{Automatic Normalization and Image Matching and Anatomical
Labeling (\animal{})}

The \animal{} algorithm carries out image registration in two phases. In the
first, a 12-parameter linear transformation (3 translations, rotations, scales,
shears) is estimated between images using an algorithm that maximizes the
correlation between blurred MR intensities and gradient magnitude over the whole
brain \citep{Collins}. In the second phase, nonlinear registration is completed
using the \animal{} algorithm \citep{Collins1995}: an iterative procedure that
estimates a 3D deformation field between two MR images. At first, large
deformations are estimated using a blurred version of the input data. These larger
deformations are then input to subsequent steps where the fit is refined by
estimating smaller deformations on data blurred with a Gaussian kernel with a
smaller full width at half maximum (FWHM). The final transformation is a set of
local translations defined on a bed of equally spaced nodes that were estimated
through the optimization of the correlation coefficient.  For the purposes of
this work we used the regularization parameters optimized in
\citet{Robbins2004}, displayed in Table \ref{tab:ANIMAL-params}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ANIMAL parameters                    
%                                     
\begin{table*}[!tbp]
\scriptsize
\caption{\textbf{\animal{} registration parameters.} \label{tab:ANIMAL-params}}
\begin{center}
\begin{tabular}{l | c c c}
\hline 
Parameters        & Stage 1 & Stage 2 & Stage 3  \tabularnewline
\hline 
Model Blur (FWHM) &   8     &    8    &   4      \tabularnewline
Input Blur (FWHM) &   8     &    8    &   4      \tabularnewline
Iterations        &   30    &    30   &   10     \tabularnewline
Step              &   8x8x8 &  4x4x4  &   2x2x2  \tabularnewline
Sub-Lattice       &   6     &    6    &   6      \tabularnewline
Lattice Diameter  &24x24x24 &12x12x12 & 6x6x6    \tabularnewline
\hline
\end{tabular}
\end{center}
\end{table*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subparagraph{Automatic Normalization Tools (\ants{})}

\ants{} is a diffeomorphic registration algorithm which provides great flexibility
over the choice of transformation model, objective function, and the consistency
of the final transformation \citep{Avants2008}. The transformation is estimated in a
hierarchical fashion where the MRI data is subsampled, allowing large
deformations to be estimated and successively refined at later hierarchical
stages (where the data is subsampled to a finer grid). The deformation field and
the objective function are regularized with a Gaussian kernel at each level of
the hierarchy. The \ants{} algorithm is freely available
\url{http://www.picsl.upenn.edu/ANTS/}. We used an implementation of the \ants{}
algorithm compatible with the MINC data format, mincANTS
\url{https://github.com/vfonov/mincANTS}.

We used the following command line when running \ants{}:
{\small
\begin{verbatim}
  mincANTS 3 -m PR[target_file.mnc,source_file.mnc,1,4] 
   --number-of-affine-iterations 10000x10000x10000x10000x10000 
   --affine-gradient-descent-option 0.5x0.95x1.e-4x1.e-4
   --use-Histogram-Matching --MI-option 32x16000
   -r Gauss[3,0] -t SyN[0.5] -i 100x100x100x20
   -o transformation.xfm
 \end{verbatim}
}
These settings were adapted from the "reasonable starting point" given in the
\ants{} manual 
\footnote{\url{https://sourceforge.net/projects/advants/files/Documentation/}}.

@@

  The {\tt xcorr\_vol} utility from the \animal{} toolkit is used to calculate the
  cross-correlation similarity measure.

@@

The automatically produces labels are compared to the \gs{} SNT labels provided by
ADNI. The segmentation accuracy reported for a parameter setting is the average
over ten validation rounds.

In order to investigate the performance of \mb{} in a real world setting in which 
only one size of atlas and template libraries are used, we explore the
variability in label agreement at fixed parameter settings under varying  
atlas and template library compositions. This is achieved by first computing the
standard deviation and variance of DSC scores in each block of ten validation
rounds per subject. The distribution of these statistics across all subjects is
then compared between \mb{} and multi-atlas using a Student's t-test. A
significant difference between distributions is taken to show either a larger or
smaller level of variability between methods.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%        Experiment 2: Hippocampal Subfield Cross-Validation          %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Experiment 2: Hippocampal Subfield Cross-Validation}
In this experiment, the accuracy of the \mb{} algorithm on hippocampal subregion
segmentation is tested using a leave-one-out cross-validation (LOOCV) design on
two image sets. The optimal parameter settings for \mb{} found in Experiment 1 are
used in this experiment.

@@

In part 2a of this experiment, we have modified the LOOCV experimental
design in order to compare \mb{} segmentation accuracy with the accuracy of image
resampling error. The five $0.3mm$-isotropic voxel Winterburn atlas images and
labels are used as the atlas library, and LOOCV is conducted using the five Winterburn
atlas images subsampled (using trilinear interpolation) to
0.9mm-isotropic voxel resolution (referred to as the {\em Subsampled} dataset)
as input subjects. Subsampling the target images allows us to assess \mb{} in a
typical segmentation scenario (high-resolution atlases and lower-resolution
subjects).  The template library is composed of all five subject images, plus
an additional set of 16 of healthy subjects 3T T1 images acquired separately
(Table \ref{tab:WAval-healthy-demographics}).  Images were acquired on a 3T GE
Discovery MR 750 system (General Electric, Milwaukee, WI) using an 8-channel
head coil with the enhanced fast gradient recalled echo 3-dimensional
acquisition protocol, EFGRE-BRAVO, with the following parameters: $TE/TR/TI =
3.0ms/6.7ms/650ms$, flip angle=$8^\circ$ , $FOV = 15.3cm$, slice
thickness$=0.9mm$, 170 in-plane steps for an approximate isotropic resolution
of 0.9mm dimension voxels.
